### 2022-03-22 omicron-generation.tex
done

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
%% More specifically, we model the logged weekly numbers of cases caused by each variant as a function of time using a shrinkage version of a cubic spline with restricted maximum likelihood.
%% JD: clarified previous s. Don't think we need this, with the cite.
Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.

@@

The generation-interval distribution for a symptom-based cohort describing \eref{serial} is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
\begin{align}
f_{G, \mathrm{symptom}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) b_I(x) \dx,
\end{align}
where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time, \fref{diagram}B):
\begin{align}
f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
\label{eq:meangen}
\end{align}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figure_epidemic.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands.}
(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates.
\label{fig:epidemic}
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_main.pdf}
\caption{
\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants.}
(A, E) Observed and fitted forward serial-interval distributions for within-household (A) and between-household (E) transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
Serial intervals are calculated for infectors who developed symptoms on weeks 50 and 51 (13--26 December, 2021).
Points represent the observed data.
Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant.
(B, F) Estimated forward generation-interval distributions for within-household (B) and between-household (F) transmission pairs in the Netherlands.
(C, D, G, H) Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta (C, G) and Omicron variants (G, H) for within-household (C, D) and between-household (G, H) transmission pairs.
Lines represent median estimates.
Shaded regions represent the corresponding 95\% confidence intervals.
\label{fig:serial}
}
\end{figure}

Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
We estimate the reproduction advantage stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore is transmission advantage (\fref{reproduction}B).
In this case, the reproduction ratio decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
Using between-household generation intervals also gives similar conclusions about changes and biases in the reproduction number estimates (Supplementary Figure S3).

@@

%A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.

%\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint from population-level epidemic dynamics (rather than individual-level transmission data).

To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.

@@

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
More generally, the bias in inferring the reproduction advantage an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate that the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.
We also show that both growth-rate differences and the reproduction advantage decreased over time---this result further corroborates are earlier theoretical framework, which showed that growth-rate differences changes over the course of an epidemic even when the reproduction advantage remain constant \citep{park2021roles}.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_rho.pdf}
\caption{
\textbf{Sensitivity of the estimates of the mean generation interval to the assumed values of the correlation coefficient of the lognormal distribution.}
Lines represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51.
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
\caption{
\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
See Figure 4 in the main text for figure caption.
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_reproduction_advantage_between.pdf}
\caption{
\textbf{Estimated time-varying reproduction number advantages of the Omicron variant using between-household generation-interval distributions.}
See Figure 5 in the main text for figure caption.
}
\end{figure}

>>>>

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We then fit a generalized additive model \citep{wood2001mgcv} to to the logged weekly case estimates to obtain smooth trajectories for case time series.
Finally, we take the derivative of the predicted logged numbers of cases caused by each variant to obtain time-varying growth rate estimates.

@@

The generation-interval distribution for a symptom-based cohort ($\tau_{g, \mathrm{symp}}$ in \eref{serial}) is biased because infectors who developed symptoms at the same time will have shorter incubation periods (when the epidemic is growing) and therefore transmit earlier (\fref{diagram}A).
This symptom-based generation-interval distribution depends on the backward incubation-period distribution:
\begin{align}
f_{G, \mathrm{symptom}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) b_I(x) \dx,
\end{align}
where $f_{G|I}(\tau|x)$ represents the forward generation-interval distribution conditional on a known value of the incubation period, $x$.
Instead, the forward generation-interval distribution measured from a cohort of individuals who were infected at the time is expected to provide reliable estimates of the distribution across individuals (because their incubation-period distribution is expected to remain constant over time, \fref{diagram}B):
\begin{align}
f_{G, \mathrm{inf}}(\tau) &= \int_0^\infty f_{G|I}(\tau|x) f_I(x) \dx.
\label{eq:meangen}
\end{align}

@@

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{figure_epidemic.pdf}
\caption{
\textbf{Epidemic dynamics on the Delta and Omicron variants in the Netherlands between November 2021 and January 2022.}
(A) Daily numbers of reported COVID-19 cases in the Netherlands (points).
The solid line represents the 7-day moving average.
Data are publicly available on \url{https://data.rivm.nl/covid-19/}.
(B) Proportion of SARS-CoV-2 variants detected from the Netherlands. Data are publicly available on \url{https://www.rivm.nl/coronavirus-covid-19/virus/varianten}.
(C) Weekly numbers of COVID-19 cases caused by the Delta (black points) and Omicron (orange triangles) variants are estimated by multiplying the weekly numbers of cases (A) with the proportion of each variant (B).
Solid lines and shaded areas represent fitted lines and corresponding 95\% confidence intervals using generalized additive model.
(D) Estimated growth rates of the Delta (black) and Omicron variants (orange) and their growth-rate differences (purple).
Lines and shaded areas represent medians and corresponding 95\% confidence intervals.
Growth rates are estimated by taking the derivative of the generalized additive model estimates.
\label{fig:epidemic}
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_main.pdf}
\caption{
\textbf{Estimated forward generation-interval distributions of Delta and Omicron variants.}
(A, E) Observed and fitted forward serial-interval distributions for within-household (A) and between-household (E) transmission pairs in the Netherlands for the Delta (black) and Omicron (orange) variants \citep{backer2021omicron}.
Serial intervals are calculated for infectors who developed symptoms on weeks 50 and 51 (13--26 December, 2021).
Points represent the observed data.
Lines represent the fitted lines assuming $r=-0.05/\textrm{day}$ for the Delta variant and $r=0.15/\textrm{day}$ for the Omicron variant.
(B, F) Estimated forward generation-interval distributions for within-household (B) and between-household (F) transmission pairs in the Netherlands.
(C, D, G, H) Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta (C, G) and Omicron variants (G, H) for within-household (C, D) and between-household (G, H) transmission pairs.
Lines represent maximum likelihood estimates.
Shaded regions represent the corresponding 95\% confidence intervals.
\label{fig:serial}
}
\end{figure}

Accounting for differences in the generation-interval distributions, we estimate that the reproduction number of the Omicron variant decreased from 1.73 (95\% CI: 1.59--1.89) to 1.14 (95\% CI: 1.00--1.32) between December 12, 2021, and January 23, 2022(\fref{reproduction}A).
On the other hand, the reproduction number of the Delta variant decreased from 0.90 (95\% CI: 0.83--0.97) to 0.69 (95\% CI: 0.65--0.75) between December 5, 2021, and January 9, 2022, and increased back up to 0.83 (95\% CI: 0.73--0.94) by January 23, 2022 (\fref{reproduction}A).
We estimate the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants) stayed roughly constant at around 2.10 (95\% CI: 1.90--2.33) between December 12--26, 2021, and slowly decreased to 1.38 (95\% CI: 1.15--1.65).
However, if we neglect differences in the generation-interval distributions and solely rely on the generation-interval-distribution estimate for the Delta variant, we over-estimate the reproduction number of the Omicron variant and therefore the reproduction advantage (\fref{reproduction}B).
In this case, the reproduction advantage decreases from 2.38 (95\% CI: 2.13--2.67) to 1.43 (95\% CI: 1.17--1.75), corresponding to roughly 4--13\% bias.
Using between-household generation intervals also gives similar conclusions about changes and biases in the reproduction number estimates (Supplementary Figure S3).

@@

A few studies have estimated that the Omicron variant has shorter transmission intervals than the Delta variant \citep{abbott2022test,kremer2022observed,song2022serial}, but there has been a lack of direct generation-interval estimates.
\cite{ito2022estimating,selby2022generation} tried to estimate the generation-interval distributions of the Omicron varaint but they both relied on population-level epidemic dynamics (rather than individual-level transmission data).
To our knowledge, our study is the first to estimate the generation-interval distribution of the Omicron variant from serial-interval data.
Although we estimate a shorter mean generation interval for the Omicron variant, we find the generation-interval distribution of the Omicron and Delta variants have similar modes (around 2.5 days), implying that the realized transmissibility of the Omicron variant decays faster.
We tentatively hypothesize that these differences may be primarily driven by the network effect \citep{park2020inferring,hart2022generation}: a higher reproduction numbers of the Omicron variant leads to faster susceptible depletion among close contacts, which in turn prevents long generation intervals from generating infections. 
While the network effect is expected to be strongest among household contacts, it is also applicable to other forms of contact structures that involve repeated contacts between the same group of individuals (because only the first infectious contact results in infection).
The network effect may also explain a decrease in the mean generation interval between week 50 and 51, especially among household transmission pairs, as a higher proportion of individuals within households would have been infected with either the Delta or Omicron variants.
Shorter generation-interval estimates for between-household contacts may be attributable to behavioral effects: individuals who have symptoms or tested positive may be more likely to stay home, preventing long between-household transmission.
Other factors, such as more stringent intervention measures against the Omicron variant \citep{backer2021omicron} and faster within-host clearance of the Omicron variant \citep{hay2022viral}, also likely contributed to shortening of generation intervals.

@@

Our study also has important implications for estimating transmission advantages of new SARS-CoV-2 variants.
In the example we consider, neglecting differences in the generation-interval distributions leads to $\approx 10\%$ bias in the estimates of the reproduction advantage (i.e., the ratio between the reproduction numbers of the Omicron and Delta variants).
More generally, the bias in inferring the reproduction advantage an emerging variant is expected to be sensitive to the assumed generation-interval distribution of the resident variant.
For example, \cite{pearson2021bounding} estimated a much higher reproduction advantage of the Omicron variant ($> 4$ fold) compared to the Delta variant in South Africa but also assumed a longer mean generation interval for the Delta and Omicron variants (6.4 vs 5.2 days, respectively).
With our generation-interval estimates, we estimate that the reproduction advantage of 2.6 for the Omicron variant assuming $r=-0.06$ and $r=0.26$ for the Delta and Omicron variants, respectively---these growth rates were chosen to match the 4-fold reproduction advantage with the estimated growth-rate differences of $0.32/\mathrm{day}$ for the Gauteng province, South Africa \cite{pearson2021bounding}.

We considered two ways of measuring transmission advantanges: growth-rate differences and reproduction advantage.
Characterizing new variants in terms of their reproduction advantage is useful because it is directly related to the amount of increased transmissibility and immune evasion \citep{pearson2021bounding}.
On the other hand, the growth-rate difference likely provides a better measure for transmission advantage for an ongoing epidemic as it provides information about which strain is growing faster.
For example, when two strains have the same $\RR$, the one with shorter generation intervals will grow faster and replace the other strain---this transmission advantage is captured by the growth-rate difference, but not by the ratio of reproduction numbers of two strains.
Therefore, we suggest using growth-rate differences and reproduction advantage as complementary measures for understanding the dynamics of emerging SARS-CoV-2 variants.

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_rho.pdf}
\caption{
\textbf{Sensitivity of the estimates of the mean generation interval to the assumed values of the correlation coefficient of the lognormal distribution.}
Lines and shaded regions represent maximum likelihood estimates and the corresponding 95\% confidence intervals for the Delta (black, solid lines) and Omicron variants (orange, dashed lines).
For illustrative purposes we use within-household serial-interval data from the cohort of infectors who developed symptoms during weeks 50 and 51 of 2021.
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_compare_stratified.pdf}
\caption{
\textbf{Estimated mean forward generation intervals of Delta and Omicron variants across different stratifications.}
Sensitivity of the mean forward generation-interval estimates to assumed growth rates of the Delta and Omicron variants stratified by the types of transmission (within- vs between-househol transmission) and the week of infectors' symptom onset (weeks 50 vs 51 of 2021).
Lines represent maximum likelihood estimates.
Shaded regions represent the corresponding 95\% confidence intervals.
}
\end{figure}

@@

\begin{figure}[!th]
\includegraphics[width=\textwidth]{figure_reproduction_advantage_between.pdf}
\caption{
\textbf{Estimated time-varying reproduction number advantages of the Omicron variant using between-household generation-interval distributions.}
(A) Estimated instantaneous reproduction numbers and their ratios over time while accounting for differences in the generation-interval distributions.
(B) Estimated instantaneous reproduction numbers and their ratios over time while assuming identical generation-interval distributions.
The instantaneous reproduction number of each variant is estimated using the renewal equation by shifting the smoothed case curves by one week (\fref{epidemic}C).
The intrinsic generation-interval distribution is approximated by the maximum likelihood estimates of the forward generation-interval distributions for between-household transmission pairs based on $r=-0.05$ for the Delta variant (black) and $r=0.15$ for the Omicron variant (orange).
Purple lines represent the ratio between the effective reproduction numbers of the Delta and Omicron variants.
Lines and shaded regions represent medians and corresponding 95\% confidence intervals.
}
\end{figure}
